---
title: "CS625-HW6"
author: "Anuradha Mantena"
date: "11/06/2020"
output: rmarkdown::github_document
---

```{r}
library(nycflights13)
library(tidyverse)
```

###### Section 5.6.7

#### Q1. Look at the number of canceled flights per day. Is there a pattern? Is the proportion of canceled flights related to the average delay?

##### Solution: 

There are multiple ways to calculate the canceled flights, one is by checking arrival delay, another is departure delay and the other is by checking the air time. We will calculate the canceled flights first. Canceled flights per day are the number of canceled flights increases with the total number of flights per day. 

```{r}
library(ggplot2)
library(dplyr)
library(tidyverse)

flights %>%
  group_by(day) %>%
  mutate(cancelled_flights = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_flights = sum(cancelled_flights),
    flights_num = n(),
  ) %>%

ggplot(aes(y =cancelled_flights, x = flights_num)) +
  geom_point()
 
```

The above figure shows the number of flights canceled increases with the total number of flights. The second part of the question asks to calculate the proportion of canceled flights related to the average delay. We can do this by finding the average departure delay or arrival delay.

```{r}
flights %>%
  group_by(day) %>%
  mutate(cancelled_flights = (is.na(arr_delay) | is.na(dep_delay))) %>%
  summarise(prop_cancelled = sum(is.na(air_time)) / n(),
          avgerage_dep_delay = mean(dep_delay, na.rm = T)) %>%
 ggplot(aes(y =prop_cancelled, x = avgerage_dep_delay)) +
  geom_point()+
   geom_smooth(method = "lm", se = FALSE)
```

We can plot the above graph by considering arr_delay, dep_delay and air_time. From the figure, it is observed that there is a strong relationship between the number of canceled flights and the average delay. The proportion of canceled flights increases with the average delay of flights.

###### Section 5.7.1

#### Q2. Which plane (tailnum) has the worst on-time record?

##### Solution:
 
To calculate the worst time there are two ways one is by checking the delay in departure which also leads to arrival delay or another way is by directly checking the arrival delay.
I grouped tailnum and then I filtered out the arr_delay column and calculated the mean for average arrival delay.

The below table shows us the information about flights along with the average delay.

```{r}
arrange(flights, desc(arr_delay)) %>%
  filter(!is.na(arr_delay)) %>%
  group_by(tailnum) %>%
  summarise(mean_arr= mean(arr_delay, na.rm =T)) %>%
  arrange(desc(mean_arr))
```

From the above results, we could see that the flight N844MH has more delaytime and it has worst on-time record followed by N911DA with 294 minutes in delay.

#### Q3. For each destination, compute the total minutes of delay. Don't need to answer the 2nd part of the question in the book.

##### Solution: 

We can answer this question in two ways, either by calculating the arrival delay or by using departure delay. Here I grouped by destination and then filtered out all the arr_delay variables. Calculate the mean for arr_delay and use arrange variable to show it in ascending order.

```{r}
flights %>% 
        group_by(dest) %>% 
        filter(!is.na(arr_delay)) %>%
  summarise(tot_delay_in_mins = sum(arr_delay[arr_delay > 0])) %>%
 arrange(tot_delay_in_mins)
```

From the above table, we could see that the LEX destination has 0 mins of delay followed by PSP destination with 36 mins of delay.


###### Section 7.3.4

#### Q4.Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)

##### Solution: 
Firstly, I wanted to check the columns in the diamonds data, I have printed all the columns. We could see there are 10 columns and 53,940 rows.

```{r}
diamonds
```
I have considered the histogram by taking a price on the x-axis. To find out the distribution of price, I considered various bin widths. In this figure, I have taken bandwidth as 10 and we could see that the count values have a range of [0-100, 500-1000, and so on..] also we can see that it has more spikes. There is a blank space in the distribution around 0 to $2500 of the price that means no diamonds were sold.

```{r}
ggplot(diamonds, aes(x = price)) +
  geom_histogram(binwidth = 10)+
  ggtitle(label = "Distribution of price")
```

From the below figures, it is observed that the count range on the y-axis varies, and also the last digits of prices are often not uniformly distributed. We could see the distribution without any blank space with the binwidth of 200. 

```{r}
ggplot(diamonds, aes(x = price)) +
  geom_histogram(binwidth = 80)+
  ggtitle(label = "Distribution of price")

ggplot(diamonds, aes(x = price)) +
  geom_histogram(binwidth = 200)+
  ggtitle(label = "Distribution of price")
```


#### Q5.How many diamonds are 0.99 carats? How many are 1 carat? What do you think is the cause of the difference?

##### Solution:

Here I have calculated the carat count by filtering only  >=0.99 'and' <=1 carat values. There is a total of 23 numbers of diamonds which are 0.99 carats and 1558 numbers of diamonds which are 1 carat.


```{r}
diamonds  %>%
  filter(carat >= 0.99 & carat <= 1) %>%
  count(carat)
```

Carat is nothing but the weight of the diamond. As cut gives the quality of the cut for diamonds I think it plays an important role when choosing the carats. Another factor is the price, the diamonds with a good price and with good cut have more count i.e 1 carat compared to 0.99 carats. I noticed here most people are interested in buying diamonds with a good cut and high carats at a reasonable price.

###### Section 7.5.1.1

#### Q6.Use what you’ve learned to improve the visualization of the departure times of canceled vs. non-canceled flights from the end of Section 7.4.

##### Solution:

From section 7.4,(see reference 1) freqpoly is used. In this section, we could see the distribution of non-canceled flights looks flat when plotted simultaneously with non-canceled flights. Using the same code I considered geom_boxplot. Here is.na() variable gives the canceled and non-canceled flight information. True represents canceled flights and False represents non-canceled flights.

```{r}
flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>%
  ggplot() +
  geom_boxplot(mapping = aes(y = sched_dep_time, x = cancelled))+
  ggtitle(label = "Canceled Vs Non-canceled flights")
```

From the above figure, it is observed that the median is high for non canceled flights and also we could see there are few outliers for true (canceled flights). Hence there are more non-canceled flights than the canceled flights. 

#### Q7.What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with a cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

##### Solution:

To find the most important variable for predicting the price of a diamond, we have to show the correlation. Here everything (): Matches all variables. we can find this by  "?select" helper for more details on how to use select variables. 

From reference 2, I have used as.numeric variable to convert all categorical values to numeric values. I have followed references 3 & 4 (see references 3 & 4) to find the correlation matrix between all the variables.
 
```{r}
diamonds %>%
  mutate(cut = as.numeric(cut),
         color = as.numeric(color),
         clarity = as.numeric(clarity)) %>%
  select(price, everything()) %>%
  cor()
```

From the above correlation, we could see that carat is the most important variable to find the price of a diamond. Here I am using Boxplot to show the relation we could see here there is a lot of variability in the distribution of carat sizes within each cut category. We could see there are outliers for the cut element "fair". 

Carat and cut are slightly negatively correlated, means diamonds of higher weights tend to have a lower cut rating Hence it shows that higher cut quality leads to lower price. 

```{r}
ggplot(diamonds, aes(x = cut, y = carat)) +
  geom_boxplot()
```

#### Extra credit (1 point): Install the ggstance package, and create a horizontal boxplot. How does this compare to using coord_flip()?

##### Solution: 

I have installed ggstance library on R. Using the coord_flip(), the horizontal plot is plotted.

```{r}
library(ggstance)

ggplot(diamonds) + 
  geom_boxplot(aes(x = cut, y = carat)) + coord_flip()
```

Without using coord_flip() also We can plot a horizontal box plot, Use "boxploth"(see reference 5) to get a horizontal boxplot. When using geom_boxploth() from ggstance, the x and y in mapping need to be switched.

```{r}
ggplot(diamonds) +
  geom_boxploth(aes(x = carat, y = cut))
```

There is no difference in the two plots, Seems like the result is the same. I could say boxploth and coord_flip() both are similar and can be used to plot for horizontal box plots.


###### Section 7.5.3.1

#### Q8.Visualize the distribution of carat, partitioned by price.

##### Solution:

We can plot this by using geom_freqpoly() and partitioning by price with cut_width. We could see that diamonds of a higher carat are associated with a higher price.


```{r}
diamonds %>% ggplot() +
  geom_freqpoly(mapping = aes(x = carat,
                             color = cut_width(price, 5000)
                             ))+
  ggtitle(label = "Distribution of carat with respect to price")
```

I have even tried with boxplot as well with 10 bins with an equal number of observations.The result is same we can observe the diamonds of a higher carat are associated with a higher price. The price of carats in the range [0.2 to 1] is medium cost.
 
```{r}
diamonds %>% ggplot +
  geom_boxplot(mapping = aes(x = cut_number(price, 10),
                             y = carat)) +
  coord_flip()
```


#### Q9.How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you?

##### Solution: 

I have considered the box plot here to find the distribution of price. I have considered 10 bins with an equal number of observations, and the width is determined by the number of observations.

We can see clearly that the price distribution of very large diamonds is much more variable than the smaller diamonds. Factors such as cut, clarity, and color have a high influence on the price of larger diamonds.

```{r}
diamonds %>% ggplot +
  geom_boxplot(mapping = aes(x = cut_number(carat, 10),
                             y = price)) +
  coord_flip()+
  ggtitle(label = "Distribution of price")
```


#### Q10.Combine two of the techniques you’ve learned to visualize the combined distribution of cut, carat, and price.

##### Solution:

Here I used box-plot to find the distribution of cut, carat, and price. From the below figure it is observed that the median is high for the carats in [1,5] range.Also we could see the carats with [0.2, 0.5] range has less price.

```{r}
ggplot(diamonds, aes(y = price, x = cut_number(carat,3), colour = cut)) +
  geom_boxplot()
```
Facetted geom_bin2d():

Another plot I have used is geom_bind2d() plot I have referred Section 7.5.3 from the document.
I used facet grid to show the all the elements in cut variable(see the reference 6). From the figure it is observed that there are more diamonds of 1 carat with price 20,000 are very good.


```{r}
diamonds %>% ggplot() +
  geom_bin2d(mapping = aes(x = carat, y = price)) +
  facet_grid(cut~.)
```

#### References:

1. Cancelled vs non -canceled flights section 7.4: https://r4ds.had.co.nz/exploratory-data-analysis.html#missing-values-2

2. Numeric conversion:
https://stackoverflow.com/questions/47922184/convert-categorical-variables-to-numeric-in-r/47923178

3. Correlation 1: http://www.sthda.com/english/wiki/correlation-matrix-a-quick-start-guide-to-analyze-format-and-visualize-a-correlation-matrix-using-r-software

4. Correlation 2:
https://stackoverflow.com/questions/5446426/calculate-correlation-for-more-than-two-variables

5. Horizontal Box plot:
https://www.rdocumentation.org/packages/ggstance/versions/0.0.0.9000/topics/geom_boxploth

6. Facet Grid:
https://stackoverflow.com/questions/22011980/map-with-geom-bin2d-overlay-with-additional-stat-info?rq=1